Staged Abstract Interpreters

### 2019-08-16 paper/10_intro.tex
Mention 2-level semantics earlier
specialization comment

Partial evaluation \cite{DBLP:books/daglib/0072559} was the first proposed
approach to realize Futamura projections: it first discovers the binding-times
(\textit{static} or \textit{dynamic}) of variables, then evaluates the static
part of the program, and generates a residualized program that solely relies on
the dynamic part. However, it is hard to precisely analyze binding-times given
an arbitrary program. As an alternative approach, multi-stage programming (MSP)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} lets users explicitly program over
a two-level semantics \cite{NIELSON198859}, by manually annotating the
binding-times of variables in the program. Then, the MSP system will check
whether these annotations are consistent, and specialize the program using that
information. The staging annotations can be either syntactic (e.g., quote and
quasi-quote in MetaML/MetaOCaml) or type-based (e.g., the Lightweight Modular
Staging framework \cite{DBLP:conf/gpce/RompfO10} in Scala).

Similar to two-level semantics \cite{NIELSON1989117}, we found
monads provide a good abstraction to hide the detail of abstract
interpretation semantics, and allows us to fit code generation into
the semantics. We use \citet{DBLP:journals/pacmpl/DaraisLNH17}'s
abstract definitional interpreter that uses monad transformers as the
unstaged baseline.  Our proposed framework adopts the type-based
multi-stage programming from the Lightweight Modular Staging framework,
and implements the Futamura projection of a big-step abstract
interpreter for a small, stateful, higher-order language. By deriving
staged monads that can be used to do code generation, we obtain a
staged abstract interpreter that fulfills the specialization. After
staging, the generated code is specialized to the input program, and
all monadic operations are inlined and compiled down to low-level
Scala code.

>>>>

Partial evaluation \cite{DBLP:books/daglib/0072559} was the first proposed
approach to realize Futamura projections: it first discovers the binding-times
(\textit{static} or \textit{dynamic}) of variables, then evaluates the static
part of the program, and generates a residualized program that solely relies on
the dynamic part. This can be viewed as a \textit{two-level semantics}:
different parts of the original program will be executed as different programs,
possibly in an environment with different semantics.
However, it is hard to precisely analyze binding-times given an arbitrary
program. As an alternative approach, multi-stage programming (MSP)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} lets users explicitly program over
a two-level semantics \cite{NIELSON198859}, by manually annotating the
binding-times of variables in the program. Then, the MSP system will check
whether these annotations are consistent, and specialize the program using that
information. The staging annotations can be either syntactic (e.g., quote and
quasi-quote in MetaML/MetaOCaml) or type-based (e.g., the Lightweight Modular
Staging framework \cite{DBLP:conf/gpce/RompfO10} in Scala).

Similar to two-level semantics \cite{NIELSON1989117}, we found
monads provide a good abstraction to hide the detail of abstract
interpretation semantics, and allows us to fit code generation into
the semantics. We use \citet{DBLP:journals/pacmpl/DaraisLNH17}'s
abstract definitional interpreter that uses monad transformers as the
unstaged baseline.  Our proposed framework adopts the type-based
multi-stage programming from the Lightweight Modular Staging framework,
and implements the Futamura projection of a big-step abstract
interpreter for a small, stateful, higher-order language. By deriving
staged monads that can be used to do code generation, we obtain a
staged abstract interpreter which is fully specialized. After
staging, the generated code is specialized to the input program, and
all monadic operations are inlined and compiled down to low-level
Scala code.
